\documentclass{ximera}

\input{../preamble}

\title{2.8 - The Intermediate Value Theorem}

%\outcome{State the Intermediate Value Theorem including hypotheses.}
%\outcome{Sketch pictures indicating why the Intermediate Value Theorem is true, and why all hypotheses are necessary.}
%\outcome{Explain why certain points exist using the Intermediate Value Theorem.}


\begin{document}

\begin{abstract}
  Here we see a consequence of a function being continuous.
\end{abstract}
\maketitle



The \textit{Intermediate Value Theorem} should not be brushed off
lightly. Once it is understood, it may seem ``obvious,'' but
we should not underestimate its power! The Intermediate Value Theorem is an important part of many proofs that establish the field of calculus and beyond.

\begin{theorem}[Intermediate Value Theorem]\label{theorem:IVT}\index{Intermediate Value Theorem}
	If $f$ is a continuous function for all $x$ in the closed interval
	$[a,b]$ and $L$ is between $f(a)$ and $f(b)$, then there is a number
	$c$ in $[a, b]$ such that $f(c) = L$.
\end{theorem}

%% \marginnote[-1.2in]{The Intermediate Value Theorem is most frequently
%%   used when $d=0$.}  
%% \marginnote[-.7in]{For a nice proof of this theorem, see: Walk, Stephen
%%   M. \textit{The intermediate value theorem is NOT obvious---and I am
%%     going to prove it to you}. College Math. J. 42 (2011), no. 4,
%%   254--259.}
%% In Figure~\ref{figure:intermediate-value}, we see a geometric
%% interpretation of this theorem.
%%BADBAD here we have a note on the proof, see the source

If you are more of a visual person, you should imagine a continuous
function, where you know the value of the function at two endpoints,
$x=a$ and $x=b$, but you don't really know what the function does
between the points $x=a$ and $x=b$:

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=0:6, ymin=0, ymax=2.2,xmax=6,
            axis lines =left, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={1,3.597,5}, ytick={.203,1,1.679},
            xticklabels={$a$,$c$,$b$}, yticklabels={$f(a)$,$f(c)=L$,$f(b)$},
            axis on top,
          ]
          \addplot [draw=none, fill=fill1, domain=(0:7)] {1.679} \closedcycle;
          \addplot [draw=none, fill=background, domain=(0:7)] {.203} \closedcycle;
          \addplot [textColor,dashed] plot coordinates {(0,1.679) (6,1.679)};
          \addplot [textColor,dashed] plot coordinates {(0,.203) (6,.203)};
          \addplot [textColor,dashed] plot coordinates {(5,0) (5,1.679)};
          \addplot [textColor,dashed] plot coordinates {(1,0) (1,.203)};
          \addplot [textColor,dashed] plot coordinates {(3.587,0) (3.597,1)};
          \addplot [penColor2,domain=(0:6)] {1};
          \addplot [very thick,penColor, smooth,domain=(0:2.5)] {sin(deg((x - 4)/2)) + 1.2};
          \addplot [very thick,penColor, smooth,domain=(4:6)] {sin(deg((x - 4)/2)) + 1.2};
          \addplot [very thick,dashed,penColor!50!background, smooth,domain=(2.5:4)] {sin(deg((x - 4)/2)) + 1.2}; 
          \addplot [color=penColor!50!background,fill=penColor!50!background,only marks,mark=*] coordinates{(3.587,1)};  %% closed hole          
          \addplot [color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,.203)};  %% closed hole          
          \addplot [color=penColor,fill=penColor,only marks,mark=*] coordinates{(5,1.679)};  %% closed hole          
        \end{axis}
\end{tikzpicture}
%% \caption{A geometric interpretation of the Intermediate Value
%%   Theorem. The function $f(x)$ is continuous on the interval
%%   $[a,b]$. Since $L$ is in the interval $[f(a),f(b)]$, there exists a
%%   value $c$ in $[a,b]$ such that $f(c) = L$.}
%% \label{figure:intermediate-value}
\end{image}
The Intermediate Value Theorem says that, even if we don't know what the function is doing between the endpoints, every intermediate output value between $f(a)$ and $f(b)$ gets used at least once along the way.

% that despite the fact that you
%don't really know what the function is doing between the endpoints, a
%point $x=c$ exists and gives an \textit{intermediate value} for $f$.

Now, let's contrast this with a time when the conclusion of the
Intermediate Value Theorem does not hold.

\begin{question}
  Consider the following situation,
  \begin{image}
    \begin{tikzpicture}
      \begin{axis}[
          domain=0:6, ymin=0, ymax=2.2,xmax=6,
          axis lines =left, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west},
          xtick={1,5}, ytick={.2,1,1.7},
          xticklabels={$a$,$b$}, yticklabels={$f(a)$,$L$,$f(b)$},
          axis on top,
        ]
        \addplot [draw=none, fill=fill1, domain=(0:7)] {1.7} \closedcycle;
        \addplot [draw=none, fill=background, domain=(0:7)] {.2} \closedcycle;
        \addplot [textColor,dashed] plot coordinates {(0,1.7) (6,1.7)};
        \addplot [textColor,dashed] plot coordinates {(0,.2) (6,.2)};
        \addplot [textColor,dashed] plot coordinates {(5,0) (5,1.7)};
        \addplot [textColor,dashed] plot coordinates {(1,0) (1,.2)};
        \addplot [penColor2,domain=(0:6)] {1};
        \addplot [very thick,penColor] plot coordinates {(1,.2) (5,.7)};
        \addplot [color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,.2)};  %% closed hole
        \addplot [color=penColor,fill=fill1,only marks,mark=*] coordinates{(5,.7)};  %% open hole          
        \addplot [color=penColor,fill=penColor,only marks,mark=*] coordinates{(5,1.7)};  %% closed hole          
      \end{axis}
    \end{tikzpicture}
  \end{image}
  and select all that are true:
  \begin{selectAll}
    \choice[correct]{$f$ is continuous on $(a,b)$.}
    \choice[correct]{$f$ is continuous on $[a,b)$.}
    \choice{$f$ is continuous on $(a,b]$.}
    \choice{$f$ is continuous on $[a,b]$.}
    \choice{There is a point $c$ in $[a,b]$ with $f(c) = L$.}
  \end{selectAll}
  
Because the hypotheses of the Intermediate Value Theorem are not all met here, the theorem does not apply. In fact, in this particular case the conclusion of the theorem is not true, since there is at least one value of $L$ that does not get used in a point. It is the discontinuity of $f(x)$ that allows some intermediate output values to get ``skipped".
\end{question}

Building on the question above, it is not difficult to see that 
each of the hypotheses of the Intermediate Value Theorem are necessary.



Let's see the Intermediate Value Theorem in action.


\begin{example}
Prove that the function $f(x)=x^3 + 3x^2+x-2$ must have a root between $0$ and $1$.


\begin{explanation}
Since $f$ is a polynomial, we see that $f$ is continuous for all real
numbers.  Since $f(0)=\answer[given]{-2}$ and
$f(1)=\answer[given]{3}$, and $0$ is between $-2$ and $3$, by the
Intermediate Value Theorem, there is a point $c$ in the interval
$[0,1]$ such that $f(c)=\answer[given]{0}$. Thus $f(x)$ must have at least one root between $0$ and $1$, although we don't know its exact value from this work alone...
\end{explanation}
\end{example}



This example also points the way to a simple method for approximating
roots. 



\begin{example} 
Approximate a root of $f(x) =x^3 + 3x^2+x-2$ between $0$ and $1$ to
within one decimal place.

\begin{explanation} 
Again, since $f$ is a polynomial, we see that $f$ is continuous for
all real numbers. Compute
\[
\begin{array}{c|c}
  x   & f(x) \\ \hline
  0.1 & \answer[given]{-1.869}\\
  0.2 & \answer[given]{-1.672}\\
  0.3 & \answer[given]{-1.403}\\
  0.4 & \answer[given]{-1.056}\\
  0.5 & \answer[given]{-0.625}\\
  0.6 & \answer[given]{-0.104}\\
  0.7 & \answer[given]{0.513}
\end{array}
\]
By the Intermediate Value Theorem, $f$ has a root between $0.6$ and
$0.7$. Repeating the process
\[
\begin{array}{c|c}
  x   & f(x) \\ \hline
  0.61 & \answer[given]{-0.046719}\\
  0.62 & \answer[given]{0.011528}
\end{array}
\]
so by the Intermediate Value Theorem, $f$ has a root between $0.61$
and $0.62$, and based on this information we can say that the root, rounded to one decimal place, is $0.6$.
\end{explanation}
\end{example}


The Intermediate Value Theorem can also be used to show that two given curves must intersect on an interval, even if actually solving for the intersection point(s) is prohibitive.


\begin{example} 
  Prove that the functions
  \begin{align*}
    f(x) &= x^8+x^6\\
    g(x) &= e^x\sqrt[3]{x+1}
  \end{align*}
  intersect on the interval $\left[-1,0\right]$.

  To start, note that both $f$ and $g$ are continuous functions because they are built from products and compositions of known continuous functions, and
  hence $h = f-g$ is also a continuous function. Now
  \begin{align*}
    h(-1) &= f(-1) - g(-1) \\
    &= \answer[given]{2} - \answer[given]{0}\\
    &= \answer[given]{2}.
  \end{align*}
  and in a similar fashion
   \begin{align*}
     h(0) &= f(0) - g(0) \\
    &= \answer[given]{0} - \answer[given]{1}\\
    &= \answer[given]{-1}.
   \end{align*}
 Thus $h(-1)$ is positive and $h(0)$ is negative; $L=0$ is a number between them. Hence by the Intermediate Value Theorem, there is some $c$ in $[-1,0]$ such that $h(c)=f(c)-g(c)=0$. Thus at $c$ we have $f(c)=g(c)$ and we conclude that $f$ and $g$
   intersect on the interval $\left[-1,0 \right]$.
%   \begin{onlineOnly}
%   We can see this point of intersection by looking at the graphs of $y=f(x)$ and $y=g(x)$ on the given interval. 
%   \[
%   \graph[xmin=1,xmax=e,ymin=-10,ymax=10]{x^2 *x^(1/3), 2(x+1)\cos(x^(1/3))}
%   \]
%   \end{onlineOnly}
\end{example}




Now we move on to a theoretical and more subtle pair of examples:

\begin{example}
Suppose you have two cats, Roxy and Yuri. Is there a time when Roxy
and Yuri have the same amount of water in their bowls assuming:
  \begin{itemize}
  \item They start and finish drinking at the same times.
  \item Roxy starts with more water than Yuri, and leaves less water
    left in her bowl than Yuri.
  \end{itemize}

\begin{explanation} 
  To solve this problem, consider two functions:
  \begin{itemize}
    \item $W_{\mathrm{Roxy}}(t) =$ the amount of water in Roxy's bowl at time $t$.
    \item $W_{\mathrm{Yuri}}(t) =$ the amount of water in Yuri's bowl at time $t$.
  \end{itemize}
  Now if $t_\mathrm{start}$ is the time the cats start drinking and
  $t_\mathrm{finish}$ is the time the cats finish drinking. Then we have
  \[
  W_\mathrm{Roxy}(t_\mathrm{start})-W_{\mathrm{Yuri}}(t_\mathrm{start}) > 0%%BADBAD dropdown answer type wanted for < > =
  \]
  and
  \[
  W_\mathrm{Roxy}(t_\mathrm{finish})-W_{\mathrm{Yuri}}(t_\mathrm{finish}) < 0.
  \]
  If we consider the amount of water in a bowl at time $t$ to be a continuous
  function (as water is ``lapped'' up in continuous small amounts),
  \[
  W_\mathrm{Roxy}-W_{\mathrm{Yuri}}
  \]
  is a continuous function, and hence the Intermediate Value Theorem
  applies. Since $W_\mathrm{Roxy}-W_{\mathrm{Yuri}}$ is positive when
  at $t_\mathrm{start}$ and negative at $t_\mathrm{finish}$, there is
  some time $t_\mathrm{equal}$ when the value is zero, meaning
  \[
  W_\mathrm{Roxy}(t_\mathrm{equal})-  W_{\mathrm{Yuri}}(t_\mathrm{equal}) =0
  \]
  meaning there is the same amount of water in each of their bowls.
\end{explanation}
\end{example}


%And finally, an example when the Intermediate Value Theorem \textit{does not} apply.

%
%\begin{example}
%Suppose you have two cats, Roxy and Yuri. Is there a time when Roxy
%and Yuri have the same amount of dry cat food in their bowls assuming:
%  \begin{itemize}
%  \item They start and finish eating at the same times.
%  \item Roxy starts with more food than Yuri, and leaves less food
%    uneaten than Yuri.
%  \end{itemize}
%
%\begin{explanation} 
%  Here we could try the same approach as before, setting:
%  \begin{itemize}
%    \item $F_{\mathrm{Roxy}}(t) =$ the amount of dry cat food in Roxy's bowl at time $t$.
%    \item $F_{\mathrm{Yuri}}(t) =$ the amount of dry cat food in Yuri's bowl at time $t$.
%  \end{itemize}
%  However in this case, the amount of food in a bowl at time $t$ is
%  \textbf{not} a continuous function! This is because dry cat food
%  consists of discrete kibbles, and is not eaten in a continuous
%  fashion. Hence the Intermediate Value Theorem \textbf{does not
%    apply}, and we can make no definitive statements concerning the
%  question above.
%\end{explanation}
%\end{example}

For some interesting extra reading check out:
\begin{itemize}
\item \link[\textit{The intermediate value theorem is NOT obvious---and I am going to prove it to you},
  S.M.\ Walk, College Math Journal, September
  2011]{http://www.jstor.org/stable/10.4169/college.math.j.42.4.254}.
\end{itemize}

\subsection{Learning Objectives}
After completing this section, students should be able to:
\vspace{.05in}

\noindent$\bullet$ State the Intermediate Value Theorem including hypotheses.
\\$\bullet$ Sketch pictures indicating why the Intermediate Value Theorem is true, and why all hypotheses are necessary.
\\$\bullet$ Prove that certain points must exist using the Intermediate Value Theorem.
%
%\outcome{State the Intermediate Value Theorem including hypotheses.}
%\outcome{Sketch pictures indicating why the Intermediate Value Theorem is true, and why all hypotheses are necessary.}
%\outcome{Explain why certain points exist using the Intermediate Value Theorem.}

\end{document}