\documentclass{ximera}

\input{../preamble.tex}

\title{2.6 - Infinite limits and vertical asymptotes}

%\outcome{Recognize when a limit is indicating there is a vertical asymptote.}
%\outcome{Evaluate the limit as $x$ approaches a point where there is a vertical asymptote.}
%\outcome{Match graphs of functions with their equations based on vertical asymptotes.}
%\outcome{Discuss what it means for a limit to equal $\infty$.}
%\outcome{Define a vertical asymptote.}
%\outcome{Understand the relationship between limits and vertical asymptotes.}
%\outcome{Find vertical asymptotes of famous functions.}
%\outcome{Find vertical asymptotes by looking at a graph.}

\begin{document}
\begin{abstract}
We explore functions that ``shoot to infinity'' near certain points.

\end{abstract}
\maketitle

Consider the function
\[
f(x) = \frac{1}{(x+1)^2}.
\]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-2:1,
            ymax=100,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(-2:-1.1)] {1/(x+1)^2};
          \addplot [very thick, penColor, smooth, domain=(-.9:1)] {1/(x+1)^2};
          \addplot [textColor, dashed] plot coordinates {(-1,0) (-1,100)};
        \end{axis}
\end{tikzpicture}
 %\caption{A plot of $f(x)=\protect\frac{1}{(x+1)^2}$.}
%% \label{plot:1/(x+1)^2}
\end{image}
While the $\lim_{x\to -1} f(x)$ does not exist, something can still be
said to indicate the precise behavior of the graph.

\begin{definition}\label{def:inflimit}\index{infinite limit}
If $\displaystyle\lim_{x\rightarrow a}f(x)$ does not exist because $f(x)$ grows arbitrarily large as $x$ approaches $a$, we write
\[
\lim_{x\to a} f(x) = \infty
\]
and say that \dfn{the limit of $f(x)$ is equal to infinity} as $x$
goes to $a$. Equivalently, we can say that as $x\rightarrow a$, $f(x)\rightarrow\infty$.

If $\displaystyle\lim_{x\rightarrow a}f(x)$ does not exist because $f(x)$ grows arbitrarily large and is negative as $x$ approaches $a$, we write
\[
\lim_{x\to a} f(x) = -\infty
\]
and say that \dfn{the limit of $f(x)$ is equal to negative infinity}
as $x$ goes to $a$. Equivalently, we can say that as $x\rightarrow a$, $f(x)\rightarrow-\infty$. 

Note that $\displaystyle\lim_{x\rightarrow a}f(x)=\pm\infty$ means that the limit does not exist, and includes additional information about the behavior of the graph.
\end{definition}

In plain terms, $\lim_{x\to a} f(x) = \infty$ means the graph heads upward near $x=a$ and $\lim_{x\to a} f(x) = -\infty$ means the graph heads downward near $x=a$. This notation indicates that the limit does not exist, but in a specific way; we do \emph{not} use this notation for individual output values like $f(-1)$.


\begin{question}
  Which of the following are correct?
  \begin{selectAll}
    \choice[correct]{$\lim_{x\to -1} \frac{1}{(x+1)^2} = \infty$}
    \choice{$\lim_{x\to -1} \frac{1}{(x+1)^2} \to \infty$}
    \choice{$f(x) = \frac{1}{(x+1)^2}$, so $f(-1) = \infty$}
    \choice[correct]{$f(x) = \frac{1}{(x+1)^2}$, so as $x\to -1$,
      $f(x) \to \infty$}
  \end{selectAll}
\end{question}

On the other hand, consider the function 
\[
f(x) = \frac{1}{x-1}.
\]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-1:2,
            ymax=50,
            ymin=-50,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(1.02:2)] {1/(x-1)};
          \addplot [very thick, penColor, smooth, domain=(-1:.98)] {1/(x-1)};
          \addplot [textColor, dashed] plot coordinates {(1,-50) (1,50)};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{1}{x-1}$.}
%% \label{plot:1/(x-1)}
\end{image}
The two sides of the limit as $x$ approaches $1$ do not agree, so $\displaystyle\lim_{x\rightarrow 1}\frac{1}{x-1}$ does not exist (often abbreviated DNE). We might consider this a ``true DNE" because it cannot be specified to $\infty$ or $-\infty$. In this case we can still consider the one-sided
limits: $\lim_{x\to 1^+} f(x) = \infty$ and $\lim_{x\to 1^-} f(x) =
-\infty$.
\\
\\You may have noticed that infinite limits and vertical asymptotes are deeply related. In fact, with this new notation we can update our definition of a vertical asymptote.


\begin{definition}\label{def:vert asymptote}\index{asymptote!vertical}\index{vertical asymptote}
At least one of the following hold
\begin{itemize}
\item $\lim_{x\to a} f(x) = \pm\infty$,
\item $\lim_{x\to a^+} f(x) = \pm\infty$,
\item $\lim_{x\to a^-} f(x) = \pm\infty$,
\end{itemize}
if and only if the line $x=a$ is a \dfn{vertical asymptote} of $f$.
\end{definition}


\begin{example}
Find the vertical asymptotes of 
\[
f(x) = \frac{x^2-9x+14}{x^2-5x+6}.
\]

\begin{explanation}
Since $f$ is a rational function, it is continuous on its domain. So the only points where the function can possibly have a vertical asymptote are zeros of the denominator. \\
Start by factoring both the numerator and the denominator:
\[
\frac{x^2-9x+14}{x^2-5x+6} = \frac{(x-2)(x-7)}{(x-2)(x-3)}
\]
Using limits, we must investigate what happens with $ f(x)$ when $x\to 2$ and $x\to 3$, since $2$ and $3$ are the only zeros of the denominator. Write
\begin{align*}
\lim_{x\to 2} \frac{\cancel{(x-2)}(x-7)}{\cancel{(x-2)}(x-3)} &= \lim_{x\to 2} \frac{(x-7)}{(x-3)}\\
&= \frac{-5}{-1}\\
&=5.
\end{align*}
Thus we have a hole in the graph at $x=2$, not a vertical asymptote.
Now write
\begin{align*}
\lim_{x\to 3} \frac{\cancel{(x-2)}(x-7)}{\cancel{(x-2)}(x-3)} &= \lim_{x\to 3} \frac{(x-7)}{(x-3)}\\
&= \lim_{x\to 3}\frac{-4}{x-3}.
\end{align*}
Consider the one-sided limits separately. 

 When $x\to 3^+$, the quantity $
(x-3)$ is positive and approaches $0$ and the numerator is negative, therefore, 
$\lim_{x\to 3^+} f(x) = -\infty$.

 On the other hand,  when $x\to 3^-$, the quantity $
(x-3)$ is negative and approaches $0$ and the numerator is negative, therefore, 
$\lim_{x\to 3^-} f(x) = \infty$.

Because these one-sided limits don't agree, we could conclude that $\lim_{x\to 3} f(x)$ does not exist. Regardless, since at least one of them is infinite we conclude that there is a vertical asymptote at $x=3$. The graph of $f(x) = \frac{x^2-9x+14}{x^2-5x+6}$ is shown below; notice the hole at $x=2$ and the vertical asymptote at $x=3$.

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=1:4,
            ymax=50,
            ymin=-50,
            samples=100,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west}
          ]
	  \addplot [very thick, penColor, smooth, domain=(3.02:4)] {(x-7)/(x-3)};
          \addplot [very thick, penColor, smooth, domain=(1:2.98)] {(x-7)/(x-3)};
          \addplot [textColor, dashed] plot coordinates {(3,-50) (3,50)};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,5)};  %% open hole
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{x^2-9x+14}{x^2-5+6}$.}
%% \label{plot:(x^2-9x+14)/(x^2-5x+6)}
\end{image}

\end{explanation}
\end{example}

Recall that rational functions are not the only functions that can have vertical asymptotes. We can adapt our knowledge about them to other circumstances.

\begin{example}
Find the vertical asymptotes of 
\[
f(x) = \frac{3xe^x}{x^2-7x}.
\]

\begin{explanation}
Again, we start by factoring both the numerator and the denominator:
\[
\frac{3xe^x}{x^2-7x} = \frac{3xe^x}{x(x-7)}
\]
Using limits, we must investigate what happens with $ f(x)$ when $x\to 0$ and $x\to 7$, since $0$ and $7$ are the only zeros of the denominator. 
\begin{align*}
\lim_{x\to 0} \frac{3\cancel{x}e^x}{\cancel{x}(x-7)} &= \lim_{x\to 0} \frac{3e^x}{x-7}\\
&= \frac{3e^0}{0-7}\\
&=-3/7.
\end{align*}
Thus we have a hole in the graph at $x=0$ and not a vertical asymptote.
\begin{align*}
\lim_{x\to 7} \frac{3\cancel{x}e^x}{\cancel{x}(x-7)} &= \lim_{x\to 7} \frac{3e^x}{x-7}\\
\end{align*}
Since the denominator here still shrinks to $0$ as $x$ approaches $7$ and the numerator does not, the limit does not exist. Consider the one-sided limits separately. 

 When $x\to 7^+$, the quantity $
(x-7)$ is positive and approaches $0$ and the numerator is positive, therefore, 
$\lim_{x\to 7^+} f(x) = \infty$.

 On the other hand,  when $x\to 7^-$, the quantity $
(x-7)$ is negative and approaches $0$ and the numerator is positive, therefore, 
$\lim_{x\to 3^-} f(x) = -\infty$.

Again, since at least one of these one-sided limits is infinite we conclude that there is a vertical asymptote at $x=7$.

\end{explanation}
\end{example}
\subsection{Learning Objectives}
After completing this section, students should be able to:
\vspace{.05in}

\noindent$\bullet$ Recognize when a limit is indicating there is a vertical asymptote.
\\$\bullet$ Evaluate the limit as $x$ approaches a point where there is a vertical asymptote.
\\$\bullet$ Match graphs of functions with their equations based on vertical asymptotes.
\\$\bullet$ Discuss what it means for a limit to equal $\infty$.
\\$\bullet$ Define a vertical asymptote.
\\$\bullet$ Understand the relationship between limits and vertical asymptotes.
\\$\bullet$ Find vertical asymptotes of famous functions.
\\$\bullet$ Find vertical asymptotes by looking at a graph.


%\outcome{Recognize when a limit is indicating there is a vertical asymptote.}
%\outcome{Evaluate the limit as $x$ approaches a point where there is a vertical asymptote.}
%\outcome{Match graphs of functions with their equations based on vertical asymptotes.}
%\outcome{Discuss what it means for a limit to equal $\infty$.}
%\outcome{Define a vertical asymptote.}
%\outcome{Understand the relationship between limits and vertical asymptotes.}
%\outcome{Find vertical asymptotes of famous functions.}
%\outcome{Find vertical asymptotes by looking at a graph.}

\end{document}