\documentclass{ximera}

\input{preamble.tex}

\title{1.1 - Functions}
\begin{document}



Our goal in this section is to define a function and explore multiple ways to describe a function.

Before we get to functions, let's define something called a \textbf{relation}. A relation is simply a rule that describes a relationship between two types of things.

For example, let's imagine we are handwashing dishes at a restaurant and we are paid two dollars per dish. This rule is a relation. It relates the number of dishes washed to amount of money paid. If we wash 30 dishes, we receive 60 dollars. In this situation, we would describe the number of dishes washed as the \textbf{input} and the amount of money paid as the \textbf{output}. Any time there is a rule relating two things, we call it a relation.

Some relations are more complex. For example, let's imagine we have a relation whose input and output are both people. If you input a person, the output will be all of that person's children. For some people, the relation will output nothing because many people have no children. For others, it will output multiple people because some people have multiple children. 

A \textbf{function} is just a specific type of relation that prevents the situations we just mentioned.

\begin{definition}\index{function}
A \dfn{function} is a relation where for each input,
there is exactly one output.
\end{definition}

\begin{question}
  If our function is the ``position with respect to time'' of some
  object, then the input is
  \begin{multipleChoice}
    \choice{position}
    \choice[correct]{time}
    \choice{none of the above}
  \end{multipleChoice}
  and the output is
  \begin{multipleChoice}
    \choice[correct]{position}
    \choice{time}
    \choice{none of the above}
  \end{multipleChoice}
\end{question}

\begin{question}
  Which of the following are functions?
  \begin{selectAll}
    \choice{Mapping words to their definition in a dictionary.}
    \choice[correct]{Mapping the U.S.\ Government's list of social security
      numbers of living people to actual living people.}
    \choice[correct]{Mapping people to their birth date.}
    \choice{Mapping mothers to their children.}
  \end{selectAll}
  \begin{feedback}\hfil
    \begin{itemize}
    \item Since words may have more than one definition, ``relating
      words to their definition in a dictionary'' is not a function.
    \item Since every social security number corresponds exactly to
      one person, ``relating social security numbers of living people
      to actual living people'' is a function.
    \item Since every person only has one birth date, ``relating
      people to their birth date'' is a function.
      \item Since mothers can have more (or less) than one child,
        ``relating mothers to their children'' is not a function.
    \end{itemize}
  \end{feedback}
\end{question}

Whenever we talk about functions, we should explicitly state
what type of things the inputs are and what type of things the outputs
are.  In calculus, functions often define a relation from (a subset
of) the real numbers (denoted by $\RR$) to (a subset of) the real
numbers.

\begin{example}
Consider the function that maps from the real numbers to the real
numbers by taking a number and mapping it to its cube:
\begin{aligned}
1 &\mapsto 1\\
-2 &\mapsto -8\\
1.5 &\mapsto 3.375
\end{aligned}
and so on. 

While we can choose to describe this rule with words, like we did here (take the input and cube it to get the output), we would like to come up with some notation which is more concise to write down. The first step is to name our function, which we often call $f$, and then the notation to say that -2 maps to -8, for example, would be $f(-2)=-8$. Reading this out loud, we say ``$f$ of -2 is -8. The input goes into the parentheses, imagine the left side of the equation as $f$ performing its rule (cubing) on -2, and then the right hand side is the output.

Of course, sometimes we want to talk about the rule itself, not acting on a specific number. In this case, we let $x$ represent any real number without being specific (we call this a \textbf{variable}) and then the notation becomes $f(x)=x^3$. This allows us to talk about the general rule of cubing a number in a concise, symbolic way.

\end{example}

\begin{warning}
\begin{itemize}
\item The parentheses in $f(x)$ do \textbf{NOT} represent multiplication. The parentheses in something like $x^2(x+1)$, on the other hand, \textbf{DO} represent multiplication, which is why we would be allowed to make a simplification like $\dfrac{x^2(x+1)}{x+1} = x^2$, because multiplication and division undo each other.
\item In this example, we named our function $f$, and we do that in many examples. Keep in mind that for more specific or important functions, we have actual names. Later we will see functions with names like $\sin, \cos, \ln$ and those are denoted $\sin(x)$, $\cos(x)$, $\ln(x)$, but they are used in exactly the same way as $f$ is here.
\end{itemize}
\end{warning}

\begin{example}

Using the same function $f(x)=x^3$, notice that we are relating numbers. $f(1)=1, f(-2)=-8$, and so on. If we only care about talking about the input and output, we can use an alternate notation like $(1,1), (-2,-8),$ etc. where the first number is the input and the second is the output. We call this an \textbf{ordered pair}, ordered because the input must come first, and pair because it is a pair of numbers.

If we did this for every possible number pair generated by $f(x)=x^3$, we would get the \textbf{graph} of the function:

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-2:2,
            width=6in,
            height=3in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor, smooth] {x^3};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=x^3$. Here we can see that for each input (a
%%   value on the $x$-axis), there is exactly one output (a value on the
%%   $y$-axis).}
%% \label{plot:fxn x^3}
\end{image}
\end{example}


%Our next example may be a function that is new to you. It is the
%\textit{greatest integer function}.
%
%\begin{example}
%Consider the \dfn{greatest integer function}.  This function maps
%any real number $x$ to the greatest integer less than or equal to $x$.
%%\[
%%f(x) = \parbox{3in}{The function that maps any real number $x$
%%  to the greatest integer less than or equal to $x$.}
%%\]
%People sometimes write this as $f(x) = \lfloor x\rfloor$, where those
%funny symbols mean exactly the words above describing the
%function. For your viewing pleasure, here is a graph of the greatest
%integer function:
%\begin{image}
%\begin{tikzpicture}
%	\begin{axis}[
%            domain=-2:4,
%            width=6in,
%            height=3in,
%            axis lines =middle, xlabel=$x$, ylabel=$y$,
%            every axis y label/.style={at=(current axis.above origin),anchor=south},
%            every axis x label/.style={at=(current axis.right of origin),anchor=west},
%            clip=false,
%            %axis on top,
%          ]
%          \addplot [textColor, very thin, domain=(0:2.3)] {0}; % puts the axis back, axis on top clobbers our open holes
%          \addplot [textColor, very thin] plot coordinates {(0,0) (0,2)}; % puts the axis back, axis on top clobbers our open holes
%	  \addplot [very thick, penColor, domain=(-2:-1)] {-2};
%          \addplot [very thick, penColor, domain=(-1:0)] {-1};
%          \addplot [very thick, penColor, domain=(0:1)] {0};
%          \addplot [very thick, penColor, domain=(1:2)] {1};
%          \addplot [very thick, penColor, domain=(2:3)] {2};
%          \addplot [very thick, penColor, domain=(3:4)] {3};
%          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(-2,-2)};  %% closed hole          
%          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(-1,-1)};  %% closed hole          
%          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(0,0)};  %% closed hole          
%          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,1)};  %% closed hole          
%          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(2,2)};  %% closed hole  
%          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(3,3)};  %% closed hole                  
%          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(-1,-2)};  %% open hole
%          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(0,-1)};  %% open hole
%          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(1,0)};  %% open hole
%          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,1)};  %% open hole
%          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(3,2)};  %% open hole
%          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(4,3)};  %% open hole
%        \end{axis}
%\end{tikzpicture}
%%% \caption{A plot of $f(x)=\lfloor x\rfloor$. Here we can see that for each input (a
%%%   value on the $x$-axis), there is exactly one output (a value on the
%%%   $y$-axis).}
%%% \label{plot:greatest-integer fxn}
%\end{image}
%Observe that here we have multiple inputs that give
%the same output.  This is not a problem! To be a function, we
%merely need to check that for each input, there is exactly one output,
%and this condition is satisfied.
%\end{example}
%
%\begin{question}
%  Compute:
%  \[
%  \lfloor 2.4 \rfloor
%  \begin{prompt}
%    =\answer{2}
%  \end{prompt}
%  \]
%  \begin{question}
%  Compute:
%  \[
%  \lfloor -2.4 \rfloor
%  \begin{prompt}
%    =\answer{-3}
%  \end{prompt}
%  \]
%\end{question}
%\end{question}


Notice that the graph of the function described above passes the so-called
\textit{vertical line test}.

\begin{theorem}
The curve $y=f(x)$ represents $y$ as a function of $x$ if and only if every possible vertical line intersects the curve $y=f(x)$ at
no more than one point. (That is, if a vertical line can intersect the curve at two or more points, the curve is not a function.) This is called the \dfn{vertical line test}.
\end{theorem}

If any vertical line drawn at a value, say $b$, could intersect the curve more than once, then that would mean $x=b$ has two outputs associated with it - that violates the ``only one output for each input" rule required to be considered a function.

\begin{definition}
  We call the set of the inputs of a function the
  \dfn{domain}\index{domain}, and we call the set of the outputs of a
  function the \dfn{range}\index{range}.
\end{definition}

Sometimes the domain and range are the \textit{entire} set of real
numbers, denoted by $\RR$. In our next examples we show that this is
not always the case.

\begin{example}
Consider the function that maps non-negative real numbers to their
positive square root. This function can be described by the
formula
\[
f(x) = \sqrt{x}.
\]
The domain is $0\le x<\infty$, which we prefer to write as
$[0,\infty)$ in interval notation. The range is $[0,\infty)$.  Here is
    a graph of $y=f(x)$:
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            xmin=-8,xmax=8,
            ymin=-5,ymax=5,
            width=6in,
            height=3in,
            domain=0:8,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor, smooth,samples=100] {sqrt(x)};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\sqrt{x}$. Here we can see that for each
%%   input (a non-negative value on the $x$-axis), there is exactly one
%%   output (a positive value on the $y$-axis).}
%% \label{plot:sqrt fxn}
\end{image}
\end{example}

To really tease out the difference between a function and its
description, let's consider an example of a function with two
different descriptions.

\begin{example}
  Explain why $\sqrt{x^2} = |x|$.
  \begin{explanation}
    Although $\sqrt{x^2}$ may appear to simplify to just $x$, let's see
    what happens when we plug in some values.
    \[
    \begin{aligned}
    \sqrt{3^2} &= \sqrt{9}\\
    &=3,
    \end{aligned}
    \qquad\text{and}\qquad
    \begin{aligned}
      \sqrt{(-3)^2} &= \sqrt{9}\\
      &=3.
    \end{aligned}
    \]
    Let $f(x)=\sqrt{x^2}$. We see that for any negative $x$, $x=-|x|$, and, therefore, $f(x)=f(-|x|)=\sqrt{(-|x|)^2}=\sqrt{|x|^2}=\answer[given]{|x|}$.  Hence
    $\sqrt{x^2}\ne x$. Rather we see that $\sqrt{x^2} = |x|$.  The
    domain of $f(x)=\sqrt{x^2}$ is $(-\infty,\infty)$ and the range is
    $[0,\infty)$.  For your viewing pleasure we've included a graph of
      $y=f(x)$:
    \begin{image}
      \begin{tikzpicture}
	\begin{axis}[
            domain=-3:3,
            width=6in,
            height=3in,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-3,...,3},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, domain=0:3] {x};
          \addplot [very thick, penColor, domain=-3:0] {-x};
        \end{axis}
      \end{tikzpicture}
    \end{image}
  \end{explanation}
\end{example}

Finally, we will consider a function whose domain is all real numbers
except for a single point.

\begin{example}
  Are 
  \[
  f(x) = \frac{x^2 - 3x + 2}{x-2} = \frac{(x-2)(x-1)}{(x-2)}
  \]
  and
  \[
  g(x) = x-1
  \]
  the same function?
  \begin{explanation}
  Let's use a series of steps to think about this question. 
    First, what if we compare graphs? Here we see a graph of
    $f$:
    \begin{image}
      \begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
          \addplot[color=penColor,fill=background,only marks,mark=*] coordinates{(2,1)};  %% open hole
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{x^2 - 3x + 2}{x-2}$. Here we
%%   can see that for each input (any value on the $x$-axis except for
%%   $x=2$), there is exactly one output (a value on the $y$-axis).}
%% \label{plot:point undfed fxn}
\end{image}
On the other hand, here is a graph of $g$:
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-2:4,
            axis lines =middle, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={-2,...,4},
            ytick={-3,...,3},
          ]
	  \addplot [very thick, penColor, smooth] {x-1};
        \end{axis}
\end{tikzpicture}
%% \caption{A plot of $f(x)=\protect\frac{x^2 - 3x + 2}{x-2}$. Here we
%%   can see that for each input (any value on the $x$-axis except for
%%   $x=2$), there is exactly one output (a value on the $y$-axis).}
%% \label{plot:point undfed fxn}
\end{image}
Second, what if we compare the domains?  We cannot evaluate $f$ at
$x=\answer[given]{2}$.  This is where $f$ is undefined.  On the other
hand, there is no value of $x$ where we cannot evaluate $g$.  In other
words, the domain of $g$ is $(-\infty, \infty)$.

Since these two functions do not have the same graph, and they do not
have the same domain, they must not be the same function.

However, if we look at the two functions everywhere except at $x = 2$, 
we can say that $f(x)=g(x)$.  In other words, 
\[
f(x)=x-1\qquad\text{when}\qquad \text{$x \ne 2$}.
\]
From this example we see that it is critical to consider the domain and
range of a function.
  \end{explanation}
\end{example}

Learning Objectives:

The following are \textbf{facts} or \textbf{definitions} you must memorize from this section:
\begin{itemize}
\item A function is a rule such that each input has exactly one output.
\item The domain of a function is the set of all inputs.
\item The range of a function is the set of all outputs.
\end{itemize}

The following is \textbf{notation} you must be familiar with from this section:
\begin{itemize}
\item $f(x)=y$, in this notation, f is the name of the function, x is the input, and y is the output.
\end{itemize}

The following are \textbf{procedures} you must be able to perform from this section:
\begin{itemize}
\item Determine whether a given formula or graph represents a function
\item Identify the domain of a function from its graph or formula
\item Identify the range of a function from its graph or formula
\item Plot a point $(x,y)$ in the plane
\item Decide whether two functions are equal to each other
\end{itemize}


\end{document}