\documentclass{ximera}

\input{../preamble.tex}

%\outcome{Identify where a function is, and is not, continuous.}
%\outcome{Understand the connection between continuity of a function and
%  the value of a limit.}
%\outcome{Make a piecewise function continuous.}

\title{Continuity of piecewise functions}

\begin{document}
\begin{abstract}
Here we use limits to ensure piecewise functions are continuous.
\end{abstract}
\maketitle
Before we start talking about continuity of piecewise functions, let's remind ourselves of all famous functions that we found to be continuous on their domains:

power functions
\\polynomials
\\rational functions
\\exponential functions
%\begin{theorem}[Continuity of Famous Functions]\index{continuity of famous functions}\label{theorem:continuity}
%he following functions are continuous on their natural domains:
%\begin{description}

%\item[Constant function]\index{constant function}
%\item[Polynomials]\index{polynomials}
%\item[Rational functions]\index{rational functions}
%\item[Power function]\index{power function}
%\item[Exponential function]\index{exponential function} 

%\item[Logarithmic function]\index{logarithmic function} 
%\item[Trig functions]\index{trigonometric functions}  
%\item[Inverse trig functions] \index{inverse trigonometric functions}    
%\end{description}
%In essence, we are saying that the functions listed above arecontinuous wherever they are defined.

%We  proved continuity of polynomials earlier using the Sum Law, Product Law and continuity of power functions.

%We  proved continuity of rational functions earlier using the Quotient Law and  continuity of polynomials.

Note that this list will eventually also include trig functions, inverse trig functions, and logarithmic functions as well. These function types are introduced later in this course.
%We can  prove continuity of the remaining four trig functions using the Quotient Law and continuity of sine and cosine functions.\\

%Since a continuous function and its inverse have ``unbroken" graphs, it follows that an inverse of a continuous function is continuous on its domain.\\
%This implies that inverse trig functions are continuous on their domains. \\

%\end{theorem}

In this section we will use our new limit skills to revisit continuity, specifically in the context of piecewise functions. Recall that $f(x)$ is continuous at $a$ if $\displaystyle\lim_{x\rightarrow a}f(x)=f(a)$, and $f(x)$ is continuous on an interval $I$ if it is continuous at every point in $I$ (considering one-sided limits at any endpoints).
%work a couple of examples involving limits, continuity, and piecewise functions.

\begin{example}
Consider the following piecewise function
\[
f(x) = 
\begin{cases}
  \frac{x}{x-1} &\text{if $x<0$}\\
  \frac{x+1}{x^2+1} + c &\text{if $x\ge 0$}.
\end{cases}
\]
Find $c$ so that $f$ is continuous at $x=0$.
\begin{explanation}
  To find $c$ such that $f$ is continuous at $x=0$, we need to find
  $c$ such that
  \[
  \lim_{x\to 0^-} f(x) = \lim_{x\to 0^+}f(x) = f(\answer[given]{0}).
  \]
  In this case
  \begin{align*}
    \lim_{x\to 0^-} f(x) &= \lim_{x\to 0^-}\answer[given]{\frac{x}{x-1}}\\
    &= \answer[given]{\frac{0}{-1}}\\
    &=\answer[given]{0}.
  \end{align*}
  On the other hand
  \begin{align*}
    \lim_{x\to 0^+} f(x) &= \lim_{x\to 0^+}\left(\answer[given]{\frac{x+1}{x^2+1} + c}\right)\\
    &= \answer[given]{\frac{0+1}{0^2+1} + c}\\
    &= \answer[given]{1+ c}
  \end{align*}
 Notice that based on the domain partitions of $f$, $f(0)=\lim_{x\to 0^+} f(x)$. Hence for our function to be continuous, we need
  \[
  1 + c = 0\qquad\text{so}\qquad c = \answer[given]{-1}.
  \]
  Now, $\lim_{x\to 0} f(x) = f(\answer[given]{0})$, and so $f$ is continuous.
\end{explanation}
\end{example}


Consider the next, more challenging example.

\begin{example}
Consider the following piecewise defined function
\[
f(x) = 
\begin{cases}
  x+4 &\text{if $x<1$}\\
  ax^2+bx+2 &\text{if $1\le x< 3$}\\
  6x+a-b &\text{if $x\ge 3$}
\end{cases}
\]
Find $a$ and $b$ so that $f$ is continuous at both $x=1$ and $x=3$.
\begin{explanation}
This problem is more challenging because we have more
unknowns. To find $a$ and
$b$ that make $f$ continuous at $x=1$, we need to find $a$ and $b$
such that
\[
\lim_{x\to 1^-} f(x) = \lim_{x\to 1^+}f(x)=f(\answer[given]{1}).
\]
Looking at the limit from the left, we have
\begin{align*}
  \lim_{x\to 1^-} f(x) &= \lim_{x\to 1^-} \left(\answer[given]{x+4}\right) \\
  &=\answer[given]{5}.
\end{align*}
Looking at the limit from the right, we have
\begin{align*}
  \lim_{x\to 1^+} f(x) &= \lim_{x\to 1^+} \left(\answer[given]{ax^2+bx+2}\right) \\
  &= \answer[given]{a+b+2}.
\end{align*}
Hence for this function to be continuous at $x=1$, we must have that
\begin{align*}
  5 &= a+b+2\\
  3 &= \answer[given]{a+b}.
\end{align*}
Notice that based on the domain partitions of $f$, $f(1)=\lim_{x\to 1^+} f(x)$. We will encounter a second requirement that results from continuity at $x=3$. Once the two requirements are combined, we will be able to solve for the values of $a$ and $b$.

To find $a$ and $b$ that make $f$ is continuous at $x=3$, we need to
find $a$ and $b$ such that
\[
\lim_{x\to 3^-} f(x) =\lim_{x\to 3^+}f(x) =f(\answer[given]{3}).
\]
Looking at the limit from the left, we have
\begin{align*}
  \lim_{x\to 3^-} f(x) &= \lim_{x\to 3^-} \left(\answer[given]{ax^2+bx+2}\right) \\
  &=\answer[given]{9a + 3b + 2}.
\end{align*}
Looking at the limit from the right, we have
\begin{align*}
  \lim_{x\to 3^+} f(x) &= \lim_{x\to 3^+} \left(\answer[given]{6x+a-b}\right) \\
  &= \answer[given]{18+a-b}.
\end{align*}
Notice that based on the domain partitions of $f$, $f(3)=\lim_{x\to 3^+} f(x)$. Hence for this function to be continuous at $x=3$, we must have that
\begin{align*}
  9a + 3b + 2 &= 18+a-b\\
  8a + 4b -16 &= 0\\
  2a + b -4 &= 0
\end{align*}
So now we have two equations and two unknowns:
\[
 3=a+b \qquad\text{and}\qquad 2a + b -4 = 0.
 \]
 Solve the first equation for $b = 3-a$ and substitute into the second equation:
 \begin{align*}
   2a + (3-a) -4&=0 \\
    a -1&=0,
 \end{align*}
 hence
 \[
 a=\answer[given]{1}\qquad\text{and so}\qquad b =3-a=\answer[given]{2}.
 \]
 Let's check, so now plugging in values for both $a$ and $b$ we find
 \[
 f(x) = 
 \begin{cases}
   x+4 &\text{if $x<1$,}\\
   \answer[given]{x^2+2x+2} &\text{if $1\le x< 3$,}\\
   \answer[given]{6x-1} &\text{if $x\ge 3$.}
\end{cases}
 \]
 Now
 \[
 \lim_{x\to 1^-} f(x) =\lim_{x\to 1^+}f(x) =f(1) =  5,
 \]
 and
 \[
 \lim_{x\to 3^-} f(x) =\lim_{x\to 3^+}f(x) =f(3) = 17.  
 \]
 So setting $a= \answer[given]{1}$ and $b=\answer[given]{2}$ makes $f$ continuous at $x=1$ and $x=3$.
 \begin{onlineOnly}
   We can confirm our results by looking at the graph of $y=f(x)$:
   \[
   \graph[xmin=-10,xmax=10,ymin=0,ymax=30]{y=x+4\left\{x<1\right\},y=x^2+2x+2\left\{1\leq x<3\right\},y=6x-1\left\{3 \leq x\right\}} 
   \]
 \end{onlineOnly}
\end{explanation}
\end{example}

\subsection{Learning Objectives}
After completing this section, students should be able to:
\vspace{.05in}

\noindent$\bullet$ Identify where a function is, and is not, continuous.
\\$\bullet$ Understand the connection between continuity of a function and the value of a limit.
\\$\bullet$ Make a piecewise function continuous.

%\outcome{Identify where a function is, and is not, continuous.}
%\outcome{Understand the connection between continuity of a function and
%  the value of a limit.}
%\outcome{Make a piecewise function continuous.}


\end{document}