\documentclass{ximera}

\input{../preamble.tex}

\title{6.4 - The Mean Value Theorem}

\outcome{Understand the statement of the Mean Value Theorem.}
\outcome{Sketch pictures to illustrate why the Mean Value Theorem is true.}
\outcome{Determine whether Rolle's Theorem or the Mean Value Theorem can be applied.}
\outcome{Find the values guaranteed by Rolle's Theorem or the Mean Value Theorem.}
\outcome{Use the Mean Value Theorem to solve word problems.}
\outcome{Compare and contrast the Intermediate Value Theorem, Mean Value Theorem, and Rolle's Theorem.}
\outcome{Identify calculus ideas which are consequences of the Mean Value Theorem.}

\begin{document}
\begin{abstract}
  Here we see a key theorem of calculus.
\end{abstract}
\maketitle


Here are some interesting questions involving derivatives:

\begin{enumerate}
\item Suppose you toss a ball upward into the air and then catch it. Must the
  ball's  velocity have been zero at some point?
\item Suppose you drive a car on a straight stretch of highway from toll booth to
  another toll booth $30$ miles away in half of an hour. Must you have
  been driving at $60$ miles per hour at some point?
\end{enumerate}

While these problems sound very different, it turns out that the
problems are very closely related. We'll start simply:

\begin{theorem}[Rolle's Theorem]\index{Rolle's Theorem} 
Suppose that $f$ is differentiable on the interval $(a,b)$,
continuous on the interval $[a,b]$, and  that $f(a)=f(b)$.
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            xmin=0, xmax=4.5,ymin=1,ymax=5,
            axis lines =left, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={1,2,3}, xticklabels={$a$,$c$,$b$},
            ytickmin=1, ytickmax=0,
            axis on top,
          ]       
          \addplot [draw=none, fill=fill1,domain=(1:3)] {5} \closedcycle;       
	  \addplot [very thick,penColor, smooth] {-(x-2)^2+4};
          \addplot [very thick,penColor2, smooth] {4};
          \node at (axis cs:.4,2.5) [penColor] {$f$}; 
          \addplot [textColor,dashed] plot coordinates {(2,0) (2,4)};
          \addplot [textColor,dashed] plot coordinates {(1,3) (3,3)};
          \addplot[color=penColor3,fill=penColor3,only marks,mark=*] coordinates{(2,4)};  %% closed hole          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,3)};  %% closed hole          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(3,3)};  %% closed hole          
        \end{axis}
\end{tikzpicture}
%% \caption{A geometric interpretation of Rolle's Theorem.}
%% \label{figure:geoRolle}
%% \end{marginfigure}
\end{image}
Then
\[
f'(c)=0
\]
for some $c$ in the open interval $(a,b)$.
%% \begin{explanation}
%% By the Extreme Value Theorem, Theorem, we know that $f$ has a
%% maximum and minimum value on $[a,b]$.

%% If maximum and minimum both occur at the endpoints, then
%% $f(x)=f(a)=f(b)$ at every point in $[a,b]$. Hence the function is a
%% horizontal line, and it has derivative zero everywhere on
%% $(a,b)$. We may choose any $c$ at all to get $f'(c)=0$.

%% If the maximum or minimum occurs at a point $c$ with $a<c<b$, then by
%% Fermat's Theorem, $f'(c)=0$.
%% \end{explanation}
\end{theorem}
\begin{explanation}
If the function $f$ happens to be  a constant, then $f'(c)=0$ for all points $c$ in the open interval $(a,b)$.

If $f$ is not a constant,  then it must attain both global extrema in the interval $[a,b]$, by the Extreme Value Theorem.
One of these two different extrema is attained in the open interval $(a,b)$ at some point $c$, and the function $f$ has a critical point there. Since $f$ is differentiable on $(a,b)$, it follows that $f'(c)=0$.
\end{explanation}

We can now answer our first question above.

\begin{example}
Suppose you toss a ball upward into the air and then catch it at the same point where you tossed it from. Must the
ball's  velocity have been zero at some point?
\begin{explanation}
  Let $p(t)$ be the position  of the ball (on the vertical line) at time $t$. Our time interval in question will be
  \[
  [t_\mathrm{start},t_\mathrm{finish}].
  \]
  We may assume that $p$ is continuous on $ [t_\mathrm{start},t_\mathrm{finish}]$ and differentiable
  on $ (t_\mathrm{start},t_\mathrm{finish})$. Since $p(t_\mathrm{start})=p(t_\mathrm{finish})$, we may now apply Rolle's Theorem and conclude that at some time
  $c$ in the given time interval, $p'(c)=\answer[given]{0}$. Hence the velocity must be zero at
  some point.
\end{explanation}
\end{example}

Rolle's Theorem is a special case of a more general theorem.

\begin{theorem}[Mean Value Theorem]\label{thm:mvt}\index{Mean Value Theorem}
Suppose that $f$ has a derivative on the interval $(a,b)$ and is
continuous on the interval $[a,b]$.
%\begin{marginfigure}[.5in]
\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            xmin=.5, xmax=5.5,ymin=0,ymax=3.1,
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            xtick={1,2.04,5}, xticklabels={$a$,$c$,$b$},
            ytickmin=1, ytickmax=0,
            axis on top,
          ] 
          \addplot [draw=none, fill=fill1,domain=(1:5)] {3.1} \closedcycle;       
          \addplot [penColor2!40!background,very thick,dashed] plot coordinates {(1,.84+1.5) (5,1.5-.96)};        
          \addplot [textColor,dashed] plot coordinates {(2.04,0) (2.04,1.5+.89)};        
	  \addplot [very thick,penColor, smooth,domain=(1:5)] {sin(deg(x))+1.5};
          \addplot [very thick,penColor2,domain=(.5:5.5)] {-.45*(x-2.04)+.89+1.5};
          %\node at (axis cs:.4,2.5) [penColor] {$f(x)$}; 
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(1,.84+1.5)};  %% closed hole          
          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(5,-.96+1.5)};  %% closed hole          
          \addplot[color=penColor3,fill=penColor3,only marks,mark=*] coordinates{(2.04,.89+1.5)};  %% closed hole          
        \end{axis}
\end{tikzpicture}
%% \caption{A geometric interpretation of the Mean Value Theorem}
%% \label{figure:geoMVT}
%% \end{marginfigure}
\end{image}
Then
\[
f'(c)=\frac{f(b)-f(a)}{b-a}
\]
for some $c$ in $(a,b)$. 
%% \begin{explanation}
%% Let 
%% \[
%% m=\frac{f(b)-f(a)}{b-a},
%% \] 
%% and consider a new function $g(x)=f(x) - m(x-a)-f(a)$.  We know that
%% $g(x)$ has a derivative on $[a,b]$, since $g'(x)=f'(x)-m$. We can
%% compute $g(a)=f(a)- m(a-a)-f(a) =0$ and
%% \begin{align*}
%% g(b)=f(b)-m(b-a)-f(a)&=f(b)-{f(b)-f(a)\over b-a}(b-a)-f(a) \\
%% &=f(b)-(f(b)-f(a))-f(a)\\
%% &=0. 
%% \end{align*}
%% So $g(a) = g(b) = 0$. Now by Rolle's Theorem, that at some $c$,
%% \[
%% g'(c)=0\qquad\text{for some $a<c<b$}.
%% \]
%% But we know that $g'(c)=f'(c)-m$, so
%% \[
%% 0=f'(c)-m=f'(c)-\frac{f(b)-f(a)}{b-a}.
%% \]
%% Hence
%% \[
%% f'(c)=\frac{f(b)-f(a)}{b-a}.
%% \]
%% \end{explanation}
\end{theorem}
Let's consider a straightforward usage of the Mean Value Theorem.

\begin{example}
    Consider $f(x)=3x^2+2x+1$ on the interval [1,3]. Determine if there exists a point that satisfies the conclusion of the Mean Value Theorem. If so, find it.
    \begin{explanation}
        Notice that $f$ is a polynomial. Therefore it is both continuous and differentiable everywhere, so it certainly is continuous on the interval $[1,3]$ and differentiable on the interval $(1,3)$. Since the hypothesis of the Mean Value Theorem is satisfied, that means the the conclusion follows. Thus we know that there exists some $c$ in $(a,b)$ such that
        
        \[ f'(c)=\frac{f(3)-f(1)}{3-1}.\]
        Let's now find that $c$ value.

    First, we need to find the derivative, $f'(x)$. Thus, $f'(x)=\answer[given]{6x+2}$. Thus,
    \[ f'(c)=6c+2=\frac{f(3)-f(1)}{3-1}=\frac{\answer[given]{34}-6}{2}=\frac{28}{2}=\answer[given]{14}.\]

    Thus, we are solving for $c$ in the equation
    \[6c+2=14.\]

    Solving, we get $c=\answer[given]{2}$. Therefore, we have found the value of $c$ that satisfies the conclusion of the Mean Value Theorem.
    \end{explanation}
\end{example}
We can now answer our second question above.  

\begin{example}
Suppose you drive a car on a straight stretch of highway from toll booth to another toll
booth $30$ miles away in half of an hour. Must you have been driving
at $60$ miles per hour at some point?

\begin{explanation}
Let $p(t)$ be the position of the car at time $t$, and $0$ hours is the
starting time with $1/2$ hours being the final time, where $p(0)=0$, and $p(1/2)=30$.
We can assume that $p$ is continuous on $[0,1/2]$ and differentiable on
$(0,1/2)$. Now the Mean Value Theorem states that there is a time $c$ such that
\[
p'(c) = \frac{30-0}{1/2-0} = \answer[given]{60}\qquad \text{where $0<c<1/2$.}
\]
Since the derivative of position is velocity, this says that the car
must have been driving at $60$ miles per hour at some point.
\end{explanation}
\end{example}
Let's examine one more example.

\begin{example}
   Assume $f$ is a polynomial. Suppose $f(0)=-3$ and $f'(x)\leq 5$ for all values of $x$. What is the maximum value of $f(2)$?
    \begin{explanation}
        This question may not feel like it is related to the Mean Value Theorem, but it is! Let's see how we can apply it to this situation. 
        
        Since $f$ is a polynomial, we know it is continuous and differentiable everywhere. That means that the Mean Value Theorem can be applied over any interval, including the interval $[0,2]$. With that in mind, we know that 
        \[ f'(c)=\frac{f(2)-f(0)}{2-0}.\]
Notice we can fill in $f(0)=-3$ and since $f'(c)\leq 5$, we will opt to let $f(c)=5$ since we are looking for the maximum value of $f(2)$. Thus,

\[ f'(c)=5=\frac{f(2)-f(0)}{2-0}=\frac{f(2)+3}{2}.\]

Now, we can solve for $f(2)$. We have,

\[5=\frac{f(2)+3}{2}.\]

Thus,

\[f(2)+3=\answer[given]{10}.\]

So $f(2)=\answer[given]{7}$.

Thus, the maximum value of $f(2)$ is $7$.

    \end{explanation}
\end{example}
Now we will address the unthinkable: could there be a continuous
function $f$ on $[a,b]$ whose derivative is zero on $(a,b)$ that is
not constant? As we will see, the answer is ``no.''

\begin{theorem} 
If $f'(x)=0$ for all $x$ in an interval $I$, then $f(x)$ is constant
on $I$.
\begin{explanation}
Let $a< b$ be two points in $I$. Since $f$ is continuous on $[a,b]$
and differentiable on $(a,b)$, by the Mean Value Theorem we know
\[
\frac{f(b)-f(a)}{b-a} = f'(c)
\]
for some $c$ in the interval $(a,b)$. Since $f'(c)=0$ we see that
$f(b)=f(a)$. %%BADBAD I would like to ask a question here.
Moreover, since $a$ and $b$ were arbitrarily chosen,
$f(x)$ must be the constant function.
\end{explanation}
\end{theorem}



In conclusion, the Mean Value Theorem relates the function $f$ and its derivative, $f'$. Since the derivative has many interpretations, e.g. instantaneous rate of change, slope of the tangent line, velocity, it is no surprise that we can use the MVT in different contexts.

Therefore, if the the function $f$ is continuous on $[a,b]$ and differentiable on $(a,b)$, then the MVT guarantees that
\begin{enumerate}
\item there is a point $c$ in $(a,b)$ where the instantaneous rate of change of $f$ is equal to the average rate of change of $f$ over the interval $[a,b]$;

or

\item there is a point $c$ in $(a,b)$ where the slope of the tangent line to the curve $y=f(x)$ is equal to the slope of the secant line through the points $(a,f(a))$  and $(b,f(b))$; 

or

\item there is a point $c$ in the time interval $(a,b)$ where the instantaneous velocity is equal to the average velocity over the time interval $[a,b]$. Here, we assume that $f$ is the position function for the object moving along a straight line.
\end{enumerate}
\end{document}